---
title: "ERC shapefile basics"
author: "Maximilian Wisnefski"
date: '2023-05-23'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rgdal)
library(sf)
library(sfheaders)
library(raster)
library(terra)
library(broom)
library(ggplot2)
library(leaflet)
library(sp)
library(s2)
library(tidyr)
library(dplyr)
library(lwgeom)
```




## Working with Shapefile
This file is mostly just for me (Max Wisnefski) to figure out how to work with shapefiles


```{r}
#unzipping wasn't really necessary, but I'll leave these as comments anyways
#system("unzip data/ERC-range.zip")
#system("unzip data/USA_Canada_ShapefileMerge.zip")
```

the shapefile of US and Canada comes from USGS
link: https://www.sciencebase.gov/catalog/item/51bf5940e4b0eb321c798ec9 

```{r}
#spdf is short for spatial polygon dataframe
US_CAN_spdf <- readOGR(dsn = "data/USA_Canada_ShapefileMerge", layer = "USA_Canada_ShapefileMerge", verbose = TRUE)
my_spdf <- readOGR(dsn = "data_2023/ERC_shapefile_2023", layer = "junivirg")
```
^I'm not sure why I am getting these warnings. I'm using both the sf and terra packages. 


### Plotting ERC range in base R
```{r}
par(mar=c(0,0,0,0))   #changes size of image
plot(my_spdf, col="#69b3a2", bg="white", lwd=0.25, border=0 )
```

```{r}
#also plotting US and some of Canada
par(mar=c(0,0,0,0))
plot(US_CAN_spdf, col="lightblue", bg="white", lwd=0.25, border=0, xlim = c(-130, -65), ylim = c(20,50))
```




```{r}
#just looking at names for the object 
names(my_spdf)
```


### plotting in ggplot2
```{r}
#must first change into df that ggplot can read
#setting region equal to different columns in @data slot of shapefile does not seem to change output at all
#I just went with CODE
spdf_tidy <- tidy(my_spdf, regions = "CODE")
```

```{r}
#produces same thing as base R
ggplot() +
  geom_polygon(data = spdf_tidy, aes( x = long, y = lat, group = group), fill="#69b3a2", color="white") +
  theme_void() 
```




```{r}
#checking to see if the two shapefiles have the same projection
#my_spdf@proj4string@projargs #does same thing as next line
crs(my_spdf)
extent(my_spdf)

crs(US_CAN_spdf)
extent(US_CAN_spdf)
#They do not have same projection
```


```{r}
#ERC projection is mercator, must change to longlat to match US/CAN projection
#changing the projection of ERC shapefile
spdf_latlon <- spTransform(my_spdf,
                                "+proj=longlat +datum=NAD83 +no_defs")
```


```{r}
#checking if change was successful
crs(spdf_latlon)
extent(spdf_latlon)
```

### Putting ERC range on top of US/CAN map

```{r}
#creating a transparent color for ERC range, so you can still see state boundaries
mycol <- rgb(105, 179, 162, max = 255, alpha = 170, names = "color")

par(mar = c(0,0,2.5,0))
plot(US_CAN_spdf,
     main = "Map of US and Canada\nWith ERC range",
     col = "white", lwd = 0.25, border = 1, xlim = c(-130, -65), ylim = c(25,50))


# add ERC range
plot(spdf_latlon,
     col = mycol,
     add = TRUE)
```

### Mapping ERC range in leaflet

I also figured out how to do this in leaflet, but I don't think it looks as good

```{r}
#making long/lat projection of ERC into df that I can use with leaflet
new_proj_tidy <- tidy(spdf_latlon, region = "CODE")
```


```{r}
outline <- new_proj_tidy[chull(new_proj_tidy$long, new_proj_tidy$lat),]

m <- leaflet(new_proj_tidy) %>% 
  addTiles() %>% 
  setView(lng = -100.45, lat = 35.25, zoom = 4 ) %>%
  #adding a dot for each point in shapefile
  addCircles(~long, ~lat, ~100, stroke = F, color = "black", group = "ERC Range") %>%
  #adding the outline
  addPolygons(data = outline, lng = ~long, lat = ~lat,
    fill = F, weight = 2, color = "darkgreen", group = "Outline of Range") %>%
  #Layers control
  addLayersControl(
    overlayGroups = c("ERC Range", "Outline of Range"),
    options = layersControlOptions(collapsed = FALSE)
  )
m
```


## Filter bird banding data based on ERC range lat/longs

Here I use the sf::st_intersects function to filter the Bird Banding Lab (BBL) data based on the ERC range shapefile

```{r}
#note that github repo data folder only has .rds since .csv was too big
BBL <- readRDS("data_2023/BBL_data/BBL2022.rds")
#changing names of lat/lon columns for convenience
colnames(BBL)[which(names(BBL) == "lat_dd")] <- "lat"
colnames(BBL)[which(names(BBL) == "lon_dd")] <- "lon"
```

```{r}
#removing rows with NA for lat/lon columns
BBL <- BBL[!is.na(BBL$lat),]
BBL <- BBL[!is.na(BBL$lat),]
```


```{r}
#making both shapefile and BBL data into sf objects so I can comepare the lats/lons
ERC_range_sf <- st_read("data_2023/ERC_shapefile_2023/junivirg.shp", quiet = TRUE) %>%
  st_transform(crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")
BBL_sf <- st_as_sf(BBL, coords = c("lon", "lat"), crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")
```


```{r}
# scatter 100 points around
#just testing st_intersects function before I actually use it on entire data set
pts = st_jitter(st_sample(ERC_range_sf, 100), factor=0.2)
#inside will be a list of boolean vals, TRUE if point falls in range, FALSE otherwise
inside = lengths(st_intersects(pts, ERC_range_sf)) > 0
out = !inside
```


```{r}
#making sure it categorized the points correctly
plot(ERC_range_sf$geom)
plot(pts[inside], pch=1, col="blue", add=TRUE);
plot(pts[out], pch=1, col="red", add=TRUE);
```

```{r}
#applying st_intersects to BBL data
inside = lengths(st_intersects(BBL_sf, ERC_range_sf)) > 0
#out = !inside
```

^This will give you a list of TRUE/FALSE vals, use this list to filter data frame, TRUE means point falls within ERC range, FALSE means it does not

```{r}
#visual from this chunk of code is not particularly useful 
#plot(ERC_range_sf$geom)
#plot(BBL_sf$geometry[inside], pch=1, col="blue", add=TRUE);
#plot(BBL_sf$geometry[out], pch=1, col="red", add=TRUE);
```


```{r}
#using the inside list to filter the BBL data to just observations within ERC range
BBL_in_range = BBL_sf %>%
  filter(BBL_sf$geometry %in% BBL_sf$geometry[inside])
```


```{r}
#making lat/lon back into separate cols (still keeping geometry column in case it ends up being useful)
#h <- head(BBL_in_range)
latlon <- do.call(rbind, st_geometry(BBL_in_range)) %>% 
    as_tibble() %>% setNames(c("lon","lat"))
BBL_in_range <- cbind(BBL_in_range, latlon)
```



```{r}
#export as .rds
#(it's too big to push to repo as a .csv)
#saveRDS(BBL_in_range, "data_2023/BBL_data/BBL_in_range_2022.rds")
```


```{r}
#also exporting unfiltered data as .rds, so I can push it to github repo 
#saveRDS(BBL, "data_2023/BBL_data/BBL2022.rds")
```

## eBird data

```{r}
#going to repeat lat/lon filtering process with eBird data
eBird_2023 <- readRDS("data_2023/dat_effort_2023.rds")
```


```{r}
#converting eBird data into an sf object 
eBird_sf <- st_as_sf(eBird_2023, coords = c("x", "y"), crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")

```


```{r}
#applying st_intersects to eBird data
inside_eBird = lengths(st_intersects(eBird_sf, ERC_range_sf)) > 0
```


```{r}
#using the inside list to filter the eBird data to just observations within ERC range
eBird_in_range = eBird_sf %>%
  filter(eBird_sf$geometry %in% eBird_sf$geometry[inside_eBird])
```


```{r}
BBL_in_range = BBL_sf %>%
  filter(BBL_sf$geometry %in% BBL_sf$geometry[inside])
```

```{r}
#making lat/lon back into separate cols (still keeping geometry column in case it ends up being useful)
#h <- head(BBL_in_range)
#latlon_eBird <- do.call(rbind, st_geometry(eBird_in_range)) %>% 
#    as_tibble() %>% setNames(c("x","y"))
#eBird_in_range <- cbind(eBird_in_range, latlon_eBird)
```


```{r}
#splitting the geometry column back into lat col and lon col by converting sf object back to df
eBird_in_range <- sf_to_df(eBird_in_range, fill = TRUE )

#removing useless columns created by above conversion
eBird_in_range = subset(eBird_in_range, select = -c(sfg_id, point_id))
```



```{r}
#export as .rds
saveRDS(eBird_in_range, "data_2023/dat_effort_in_range_2023.rds")
```

