---
title: "ERC shapefile basics"
author: "Maximilian Wisnefski"
date: '2023-05-23'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#library(rgdal)
library(tidyverse)
library(sf)
library(mapdata)

library(sfheaders)
library(raster)
library(terra)
library(broom)
#library(leaflet)
library(sp)
library(s2)
library(lwgeom)
# library(rnaturalearth)
# library(rnaturalearthhires)
```

# Introduction
This code file imports a shapefile representing an updated range boundary for the eastern redcedar tree, Juniperus virginia. The shapefile was created by W. Haas, 2023 (Kent State University).
The file also imports occurrence data from eBird and the Breeding Bird Laboratory. 
It uses the shapefile as a filter to identify occurrences that fall within the eastern redcedar (ERC) species range.

The file was originally authored by M. Wisnefski (Anderson Summer Scholar, 2023) and updated by S. Supp (2024-2025).


## Prepare the Eastern Redcedar range data

Import the ERC range file, updated by W. Haas, 2022
```{r}
ERC_range_sf <- st_read("data_2023/ERC_shapefile_2023/junivirg.shp", quiet = TRUE) %>%
  st_transform(crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")
```

Plot the shapefile on a map of North America.
Get the map boundaries.
```{r}
#Get polygons for north america map
world_map <- map_data("world")
north_america <- world_map %>%
  filter(region %in% c("USA", "Canada"))
```

plot the range shapefile on a map to check that it's working
```{r}
# plot the number of checklists submitted (summed) per grid cell (2008-2023)
ggplot() + 
  geom_sf(data = ERC_range_sf, fill = "lightgreen", color = alpha("darkgreen")) +
  geom_polygon(data=north_america, aes(x=long, y=lat, group=group), fill=NA, color="black") +
  coord_sf(xlim = c(-103, -65), ylim = c(25, 50)) + #set limits without distorting boundaries
  labs(title = "W. Haas shapefile of J. virginiana range extent") +
  theme_minimal() + 
  theme(axis.line=element_blank(),axis.text.x=element_blank(),
          axis.text.y=element_blank(),axis.ticks=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank())

ggsave(filename = "figs/ERC_range_map.png", height = 4, width=5.5)
```


## Filter within range eBird data, 2008-2023

Read in the eBird data that was summarized in previous code
```{r}
#going to repeat lat/lon filtering process with eBird data
eBird_2023 <- readRDS("data_2023/dat_effort_2023.rds")
```

Convert the eBird dataframe into an sf object
```{r}
#converting eBird data into an sf object 
eBird_sf <- st_as_sf(eBird_2023, 
                     coords = c("x", "y"), 
                     crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")
```

Find out which avian occurrences fall within the eastern redcedar range (ERC_range_sf)
Geometry for eBird_sf represents the center point of the icosahedron cell within which the observation took place.
Geometry for ERC_range_sf represents the range boundaries for eastern red cedar tree

Note: The filter step requires several minutes to run on a desktop computer.
```{r}
# apply st_intersects to eBird data
inside_eBird = lengths(st_intersects(eBird_sf, ERC_range_sf)) > 0

# use inside_eBird to filter the eBird data to only include observations within ERC range extent
eBird_in_range = eBird_sf %>%
  filter(eBird_sf$geometry %in% eBird_sf$geometry[inside_eBird])
```

clean up the filtered data to write back to tabular format for storing
```{r}
#splitting the geometry column back into lat col and lon col by converting sf object back to df
eBird_in_range <- sf_to_df(eBird_in_range, fill = TRUE )

#removing unnecessary columns created by above conversion
eBird_in_range = subset(eBird_in_range, select = -c(sfg_id, point_id))
```

Save filtered data as a file to the main project folder
```{r}
#export as .rds
saveRDS(eBird_in_range, "data_2023/dat_effort_in_range_2023.rds")
```

```{r}
# plot the checklists submitted within the range
ggplot() + 
  geom_sf(data = ERC_range_sf, fill = "lightgreen", color = alpha("darkgreen")) +
  geom_polygon(data=north_america, aes(x=long, y=lat, group=group), fill=NA, color="black") +
  geom_point(data=eBird_in_range, aes(x, y, alpha=0.25)) +
  coord_sf(xlim = c(-103, -65), ylim = c(25, 50)) + #set limits without distorting boundaries
  labs(title = "eBird occurrences within J. virginiana range extent") +
  theme_minimal() + 
  theme(axis.line=element_blank(),axis.text.x=element_blank(),
          axis.text.y=element_blank(),axis.ticks=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank())
```




## Filter within range BBL data, 2008-2023

Read in the BBL data and conduct some basic data preparation steps
```{r}
#note that github repo data folder only has .rds since .csv was too big
BBL <- readRDS("data_2023/BBL_data/BBL2022.rds")

#changing names of lat/lon columns for convenience
colnames(BBL)[which(names(BBL) == "lat_dd")] <- "lat"
colnames(BBL)[which(names(BBL) == "lon_dd")] <- "lon"

#removing rows with NA for lat/lon columns
BBL <- BBL[!is.na(BBL$lat),]
BBL <- BBL[!is.na(BBL$lat),]
```

convert the BBL dataframe into an sf object
```{r}
BBL_sf <- st_as_sf(BBL, coords = c("lon", "lat"), crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")
```

Find out which BBL records occur within the ERC range
```{r}
#applying st_intersects to BBL data
inside = lengths(st_intersects(BBL_sf, ERC_range_sf)) > 0

BBL_in_range = BBL_sf %>%
  filter(BBL_sf$geometry %in% BBL_sf$geometry[inside])
```

clean up the filtered data to write back to tabular format for storing
```{r}
#splitting the geometry column back into lat col and lon col by converting sf object back to df
BBL_in_range <- sf_to_df(BBL_in_range, fill = TRUE )

#removing unnecessary columns created by above conversion
BBL_in_range = subset(BBL_in_range, select = -c(sfg_id, point_id))

#rename columns
BBL_in_range <- rename(BBL_in_range, lon = x)
BBL_in_range <- rename(BBL_in_range, lat = y)
```

Save filtered data as a file to the main project folder
```{r}
#export as .rds
saveRDS(BBL_in_range, "data_2023/BBL_data/BBL_in_range_2024.rds")
```

```{r}
# plot the number of records submitted within the range
ggplot() + 
  geom_sf(data = ERC_range_sf, fill = "lightgreen", color = alpha("darkgreen")) +
  geom_polygon(data=north_america, aes(x=long, y=lat, group=group), fill=NA, color="black") +
  geom_point(data=BBL_in_range, aes(lon, lat, alpha=0.25)) +
  coord_sf(xlim = c(-103, -65), ylim = c(25, 50)) + #set limits without distorting boundaries
  labs(title = "BBL records within J. virginiana range extent") +
  theme_minimal() + 
  theme(axis.line=element_blank(),axis.text.x=element_blank(),
          axis.text.y=element_blank(),axis.ticks=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank())
```




```{r}

```

```{r}
#making lat/lon back into separate cols (still keeping geometry column in case it ends up being useful)
#h <- head(BBL_in_range)
#latlon_eBird <- do.call(rbind, st_geometry(eBird_in_range)) %>% 
#    as_tibble() %>% setNames(c("x","y"))
#eBird_in_range <- cbind(eBird_in_range, latlon_eBird)
```

















## Working with Shapefile
This file is mostly just for me (Max Wisnefski) to figure out how to work with shapefiles


```{r}
#unzipping wasn't really necessary, but I'll leave these as comments anyways
#system("unzip data/ERC-range.zip")
#system("unzip data/USA_Canada_ShapefileMerge.zip")
```

the shapefile of US and Canada comes from USGS
link: https://www.sciencebase.gov/catalog/item/51bf5940e4b0eb321c798ec9 

```{r}
#spdf is short for spatial polygon dataframe
US_CAN_spdf <- readOGR(dsn = "old-code-data/data/USA_Canada_ShapefileMerge", layer = "USA_Canada_ShapefileMerge", verbose = TRUE)
my_spdf <- readOGR(dsn = "data_2023/ERC_shapefile_2023", layer = "junivirg")
```
^I'm not sure why I am getting these warnings. I'm using both the sf and terra packages. 


### Plotting ERC range in base R
```{r}
par(mar=c(0,0,0,0))   #changes size of image
plot(my_spdf, col="#69b3a2", bg="white", lwd=0.25, border=0 )
```

```{r}
#also plotting US and some of Canada
par(mar=c(0,0,0,0))
plot(US_CAN_spdf, col="lightblue", bg="white", lwd=0.25, border=0, xlim = c(-130, -65), ylim = c(20,50))
```




```{r}
#just looking at names for the object 
names(my_spdf)
```


### plotting in ggplot2
```{r}
#must first change into df that ggplot can read
#setting region equal to different columns in @data slot of shapefile does not seem to change output at all
#I just went with CODE
spdf_tidy <- tidy(my_spdf, regions = "CODE")
```

```{r}
#produces same thing as base R
ggplot() +
  geom_polygon(data = spdf_tidy, aes( x = long, y = lat, group = group), fill="#69b3a2", color="white") +
  theme_void() 
```




```{r}
#checking to see if the two shapefiles have the same projection
#my_spdf@proj4string@projargs #does same thing as next line
crs(my_spdf)
extent(my_spdf)

crs(US_CAN_spdf)
extent(US_CAN_spdf)
#They do not have same projection
```


```{r}
#ERC projection is mercator, must change to longlat to match US/CAN projection
#changing the projection of ERC shapefile
spdf_latlon <- spTransform(my_spdf,
                                "+proj=longlat +datum=NAD83 +no_defs")
```


```{r}
#checking if change was successful
crs(spdf_latlon)
extent(spdf_latlon)
```

### Putting ERC range on top of US/CAN map

```{r}
#creating a transparent color for ERC range, so you can still see state boundaries
mycol <- rgb(105, 179, 162, max = 255, alpha = 170, names = "color")

par(mar = c(0,0,2.5,0))
plot(US_CAN_spdf,
     main = "Map of US and Canada\nWith ERC range",
     col = "white", lwd = 0.25, border = 1, xlim = c(-130, -65), ylim = c(25,50))


# add ERC range
plot(spdf_latlon,
     col = mycol,
     add = TRUE)
```


### Mapping ERC range on US map with ggplot

```{r}
#making the returnclass sf, so I can zoom in on it 
state_prov <- rnaturalearth::ne_states(c("united states of america", "canada", "mexico", "honduras", "el salvador", "guatemala", "costa rica"), returnclass = "sf")
state_prov <- st_crop(state_prov, xmin = -110, xmax = -50, ymin = 20, ymax = 50)
```


```{r}
range_sf = st_as_sf(spdf_latlon, crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")

ggplot(data=state_prov, colour="grey") + 
  geom_sf(fill = "white") + 
  geom_sf(data = range_sf, fill = mycol) +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))

#ggsave("../ERC_range.png", bg = "transparent", height = 10, width = 15)
```



### Mapping ERC range in leaflet

I also figured out how to do this in leaflet, but I don't think it looks as good

```{r}
#making long/lat projection of ERC into df that I can use with leaflet
new_proj_tidy <- tidy(spdf_latlon, region = "CODE")
```


```{r}
outline <- new_proj_tidy[chull(new_proj_tidy$long, new_proj_tidy$lat),]

m <- leaflet(new_proj_tidy) %>% 
  addTiles() %>% 
  setView(lng = -100.45, lat = 35.25, zoom = 4 ) %>%
  #adding a dot for each point in shapefile
  addCircles(~long, ~lat, ~100, stroke = F, color = "black", group = "ERC Range") %>%
  #adding the outline
  addPolygons(data = outline, lng = ~long, lat = ~lat,
    fill = F, weight = 2, color = "darkgreen", group = "Outline of Range") %>%
  #Layers control
  addLayersControl(
    overlayGroups = c("ERC Range", "Outline of Range"),
    options = layersControlOptions(collapsed = FALSE)
  )
m
```


## Filter bird banding data based on ERC range lat/longs

Here I use the sf::st_intersects function to filter the Bird Banding Lab (BBL) data based on the ERC range shapefile

```{r}
#note that github repo data folder only has .rds since .csv was too big
BBL <- readRDS("data_2023/BBL_data/BBL2022.rds")
#changing names of lat/lon columns for convenience
colnames(BBL)[which(names(BBL) == "lat_dd")] <- "lat"
colnames(BBL)[which(names(BBL) == "lon_dd")] <- "lon"
```

```{r}
#removing rows with NA for lat/lon columns
BBL <- BBL[!is.na(BBL$lat),]
BBL <- BBL[!is.na(BBL$lat),]
```


```{r}
#making both shapefile and BBL data into sf objects so I can comepare the lats/lons
ERC_range_sf <- st_read("data_2023/ERC_shapefile_2023/junivirg.shp", quiet = TRUE) %>%
  st_transform(crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")
BBL_sf <- st_as_sf(BBL, coords = c("lon", "lat"), crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")
```


```{r}
# scatter 100 points around
#just testing st_intersects function before I actually use it on entire data set
pts = st_jitter(st_sample(ERC_range_sf, 100), factor=0.2)
#inside will be a list of boolean vals, TRUE if point falls in range, FALSE otherwise
inside = lengths(st_intersects(pts, ERC_range_sf)) > 0
out = !inside
```


```{r}
#making sure it categorized the points correctly
plot(ERC_range_sf$geom)
plot(pts[inside], pch=1, col="blue", add=TRUE);
plot(pts[out], pch=1, col="red", add=TRUE);
```

```{r}
#applying st_intersects to BBL data
inside = lengths(st_intersects(BBL_sf, ERC_range_sf)) > 0
#out = !inside
```

^This will give you a list of TRUE/FALSE vals, use this list to filter data frame, TRUE means point falls within ERC range, FALSE means it does not

```{r}
#visual from this chunk of code is not particularly useful 
#plot(ERC_range_sf$geom)
#plot(BBL_sf$geometry[inside], pch=1, col="blue", add=TRUE);
#plot(BBL_sf$geometry[out], pch=1, col="red", add=TRUE);
```


```{r}
#using the inside list to filter the BBL data to just observations within ERC range
BBL_in_range = BBL_sf %>%
  filter(BBL_sf$geometry %in% BBL_sf$geometry[inside])
```


```{r}
#making lat/lon back into separate cols (still keeping geometry column in case it ends up being useful)
#h <- head(BBL_in_range)
latlon <- do.call(rbind, st_geometry(BBL_in_range)) %>% 
    as_tibble() %>% setNames(c("lon","lat"))
BBL_in_range <- cbind(BBL_in_range, latlon)
```



```{r}
#export as .rds
#(it's too big to push to repo as a .csv)
#saveRDS(BBL_in_range, "data_2023/BBL_data/BBL_in_range_2022.rds")
```


```{r}
#also exporting unfiltered data as .rds, so I can push it to github repo 
#saveRDS(BBL, "data_2023/BBL_data/BBL2022.rds")
```
