---
title: "ME-paper-fig-example"
author: "Sarah Supp"
date: "4/30/2021"
output: html_document
---

#code to make examples using eBird data for the Movement Ecology paper

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(data.table)
library(devtools)
#devtools::install_github("dkahle/ggmap")
library(ggmap)
library(maptools)
library(fields)
library(sp)
library(raster)
library(rgdal)
library(rgeos)
library(sf)
library(maps)
library(mapdata)
library(mgcv)
library(gamm4) 
library(tidyverse) 
library(ggpubr)
#devtools::install_github('r-barnes/dggridR', vignette=TRUE)
library(dggridR)
library(RColorBrewer)
library(geosphere)
library(radiant.data)
library(rnaturalearth)
#devtools::install_github("ropensci/rnaturalearthhires")
library(rnaturalearthhires)
#library(SDMTools)#FIXME: package ‘SDMTools’ is not available (for R version 3.6.2)
#devtools::install_github("r-spatial/sf") #C compiler error on mac
#library(geojsonio)
```

#Making example fig for data processing from eBird

Each species has it's own merged data file, and will be brought into one species data (spdata) file to work through the results. I'll make an example figure with Yellow-rumped Warbler in 2019
```{r}
files = list.files(path=here(), recursive=T, pattern=glob2rx("dat_effort*.rds"))

#for Figure 2
yewa <- readRDS(paste0(here(), "/data/dat_effort_YEWA.rds")) 
yewa <- yewa %>% filter(YEAR==2019)

#for Figure 3
robi <- readRDS(paste0(here(), "/data/dat_effort_ROBI.rds"))
robi <- robi %>% filter(lon >= -103)
bchu <- readRDS("/home/sarah/Documents/GitHub/hb-mig-env/data/dat_effort_bchu.rds")
bchu <- bchu %>% filter(lon <= -103)

#load annas hb
load(paste0(here(), "/data/FAL-data-update/ANHU-count-effort.RData"))
anhu <- dat2 %>% select(POLYFID, YEAR=year, DAY=day, count.x, count.y) %>% mutate(species="Calypte anna")

robi_bchu <- rbind(robi, bchu)
robi_bchu <- filter(robi_bchu, YEAR != 2020) #2020 is incomplete, don't include

robi_bchu <- robi_bchu %>% select(POLYFID, YEAR, DAY, count.x, count.y, species)
robi_bchu_anhu <- rbind(robi_bchu, anhu)

#
```


1. Filter to species: plot all the points (note that these are binned points, so they might look a little strange)
```{r}
dim(yewa)

# get background map layers
coastlines <- rnaturalearth::ne_coastline()
state_prov <- rnaturalearth::ne_states(c("united states of america", "canada", "mexico", "honduras", "el salvador", "guatemala", "costa rica"))
us  <- map_data("usa")

ggplot() + 
  geom_polygon(data=coastlines, aes(x=long, y=lat, group = group), colour="grey50", fill="white") + 
  geom_point(data=yewa, aes(lon, lat), col="darkorchid4", alpha=.2, size=.1) + 
 # xlim(-102,-52) + 
  theme_bw() + 
  theme_void()
#  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
```

2. count the points and check for enough and consistency
```{r}
n = nrow(yewa)

#count total number of records across the years. (increasing strongly)
counts <- yewa %>%
  group_by(species, YEAR) %>%
  tally()

num_bins <- ggplot(counts, aes(YEAR, n)) + geom_bar(stat="identity") + 
  theme_bw() + ylab("Number of binned observations") + xlab("Year") +
  theme(axis.text.x = element_text(angle = 90)) + 
  facet_wrap(~species, ncol=1)

#checking how many unique grid cells logged an observation in each year (relatively flat)
nPOLYFID <- yewa %>%
  group_by(species, DAY) %>%
  summarise(n=n_distinct(POLYFID))

num_gridcells <- ggplot(nPOLYFID, aes(DAY, n)) + geom_bar(stat="identity") + 
  theme_bw() + ylab("Number of unique grid cells") + xlab("Day of Year") +
  theme(axis.text.x = element_text(angle = 90)) + 
  facet_wrap(~species, ncol=1)

#checking how many days per year have at least 1 observation (ALL of them)
nDAYSofyear <- yewa %>%
  group_by(species, DAY) %>%
  summarise(n=n())

num_days <- ggplot(nDAYSofyear, aes(DAY, n)) + geom_bar(stat="identity") + 
  theme_bw() + ylab("Number of unique records per day") + xlab("Day of Year") +
  theme(axis.text.x = element_text(angle = 90)) + 
  facet_wrap(~species, ncol=1)

ggarrange(num_bins, num_gridcells, num_days,
          labels=c("A", "B", "C"),
          ncol=3, nrow=1)

# make a single plot for Fig 2.
ggplot(nPOLYFID, aes(DAY, n)) + geom_bar(stat="identity") + 
  theme_bw() + ylab("Number unique grid cells") + xlab("Day of Year") +
  #theme(axis.text.x = element_text(angle = 90)) + 
  theme(text = element_text(size = 20)) +
  facet_wrap(~species, ncol=1)

# make a single plot for Fig 2.
ggplot(nPOLYFID, aes(DAY, n)) + geom_line() + 
  theme_bw() + ylab("Number unique grid cells") + xlab("Day of Year") +
  #theme(axis.text.x = element_text(angle = 90)) + 
  theme(text = element_text(size = 20)) +
  facet_wrap(~species, ncol=1)

ggsave(filename=paste0(here(),"/figs/yewa_gridcells.png"), width=5, height=4)


range(nPOLYFID$n, na.rm=TRUE)
hist(nPOLYFID$n)

```

Weight occurrence records by effort and plot
```{r}
#weights all together
weight_yewa <- yewa %>%
    mutate(DAY = as.numeric(DAY),
         YEAR = as.numeric(YEAR)) %>%
group_by(species, YEAR, DAY) %>%
  summarise(
            numcells = n(), 
            numobs = sum(count.x),
            wtmean_lon = weighted.mean(lon, count.x/count.y), 
            wtmean_lat = weighted.mean(lat, count.x/count.y), 
            wtsd_lon = weighted.sd(lon, count.x/count.y),
            wtsd_lat = weighted.sd(lat, count.x/count.y) 
            ) %>%
  ungroup()

weight_yewa <- weight_yewa %>%
  mutate(DAYm1 = DAY-1,
         origin = paste0(YEAR, "-01-01"),
         DATE = as.Date(DAYm1, origin=origin), 
         MONTH = month(DATE)) %>%
  mutate(winter = ifelse(MONTH %in% c(10, 11, 12, 1, 2, 3), "winter", "non-winter")) %>%
  arrange(species,DATE) %>%
  ungroup() %>%
  mutate(ID = row_number()) %>%
  select(-DAYm1, -origin)

## Calculates the count of observations within each cell, 
#weighted by total eBirder effort on that day in a given cell. 
#Appends into the dataframe as "count_weighted" for analysis. Weighted as the total number of observations of the #target species in a cell divided by the total number of eBirder records in a cell.

dgg <- dgconstruct(project = "FULLER", aperture = 4, topology = "HEXAGON", res = 6)
yewa$cell <- dgGEO_to_SEQNUM(dgg, yewa$lon, yewa$lat)$seqnum
yewa$cell_lat <- dgSEQNUM_to_GEO(dgg, yewa$cell)$lat_deg 
yewa$cell_lon <- dgSEQNUM_to_GEO(dgg, yewa$cell)$lon_deg 

yewa <- yewa %>%
  mutate(count_weighted = count.x/count.y)

spp_counts <- yewa %>%
  group_by(species, cell) %>% 
  summarise(sum_weighted_count=sum(count_weighted),
            mean_weighted_count=mean(count_weighted),
            sum_count=sum(count.x))

#Get the grid cell boundaries for cells which had bird observations
grid <- dgcellstogrid(dgg, spp_counts$cell, frame=TRUE, wrapcells=TRUE)

#Update the grid cells' properties to include the number of observations in each cell
grid <- merge(grid, spp_counts, by.x="cell", by.y="cell")

ggplot() + 
  geom_polygon(data=coastlines, aes(x=long, y=lat, group = group), colour="grey50", fill="white") + 
  #geom_polygon(data=countries, aes(x=long, y=lat, group=group), fill=NA, color="black")   +
  geom_polygon(data=grid, aes(x=long, y=lat, group=cell, fill=mean_weighted_count), size=0.001)    +
  geom_path   (data=grid, aes(x=long, y=lat, group=cell), alpha=0.4, color=NA) +
#  geom_point  (aes(x=cellcenters$lon_deg, y=cellcenters$lat_deg), size=0.5) +
  scale_fill_gradient(low="gray90", high="darkorchid4") + 
  #scale_fill_gradient2(low="blue", high="red", midpoint = 250) 
  theme_void() + 
  facet_wrap(~species)

```
Calculates weights separately by east vs west flyway

```{r}
#weights EASTERN flyway only
weight_yewa_east <- yewa %>%
  filter(lon>-103) %>%
    mutate(DAY = as.numeric(DAY),
         YEAR = as.numeric(YEAR)) %>%
group_by(species, YEAR, DAY) %>%
  summarise(
            numcells = n(), 
            numobs = sum(count.x),
            wtmean_lon = weighted.mean(lon, count.x/count.y), 
            wtmean_lat = weighted.mean(lat, count.x/count.y), 
            wtsd_lon = weighted.sd(lon, count.x/count.y),
            wtsd_lat = weighted.sd(lat, count.x/count.y) 
            ) %>%
  ungroup() %>%
  mutate(DAYm1 = DAY-1,
         origin = paste0(YEAR, "-01-01"),
         DATE = as.Date(DAYm1, origin=origin), 
         MONTH = month(DATE)) %>%
  mutate(winter = ifelse(MONTH %in% c(10, 11, 12, 1, 2, 3), "winter", "non-winter")) %>%
  arrange(species,DATE) %>%
  ungroup() %>%
  mutate(ID = row_number()) %>%
  select(-DAYm1, -origin)

#weights WESTERN flyway only
weight_yewa_west <- yewa %>%
  filter(lon<=-103) %>%
    mutate(DAY = as.numeric(DAY),
         YEAR = as.numeric(YEAR)) %>%
group_by(species, YEAR, DAY) %>%
  summarise(
            numcells = n(), 
            numobs = sum(count.x),
            wtmean_lon = weighted.mean(lon, count.x/count.y), 
            wtmean_lat = weighted.mean(lat, count.x/count.y), 
            wtsd_lon = weighted.sd(lon, count.x/count.y),
            wtsd_lat = weighted.sd(lat, count.x/count.y) 
            ) %>%
  ungroup() %>%
  mutate(DAYm1 = DAY-1,
         origin = paste0(YEAR, "-01-01"),
         DATE = as.Date(DAYm1, origin=origin), 
         MONTH = month(DATE)) %>%
  mutate(winter = ifelse(MONTH %in% c(10, 11, 12, 1, 2, 3), "winter", "non-winter")) %>%
  arrange(species,DATE) %>%
  ungroup() %>%
  mutate(ID = row_number()) %>%
  select(-DAYm1, -origin)
```


*Creates a dataframe with daily location for each year and graphs results*
```{r}
Estimatedailylocs = function(dat) {
  
  x <- unique(dat$YEAR)
  spp <- unique(dat$species)

  df <- data.frame(species = "none", DAY = 0, YEAR =0, MONTH = 0, winter = "not", lon = 0, lat = 0, lon_se = 0, lat_se = 0)
  
  for (s in spp){
    for (i in x) {
      sub_weighted_mean_locs <- 
        dat %>%
        filter(YEAR == i & species == s)
      
      lon_gam = gam(wtmean_lon ~ s(DAY, k=20), data = sub_weighted_mean_locs, gamma = 1.5)
      lat_gam = gam(wtmean_lat ~ s(DAY, k=20), data = sub_weighted_mean_locs, gamma = 1.5)
      xpred = data.frame(DAY=sort(unique(sub_weighted_mean_locs$DAY)))
      lonpred = predict(lon_gam, newdata = xpred, type="response", se.fit=T)
      latpred = predict(lat_gam, newdata = xpred, type="response", se.fit=T)
      
      preds =  data.frame(species = s, DAY = xpred$DAY, YEAR = i, 
                          MONTH = sub_weighted_mean_locs$MONTH, winter = sub_weighted_mean_locs$winter, 
                          lon = lonpred$fit, lat = latpred$fit, lon_se = lonpred$se.fit, lat_se = latpred$se.fit)
      df <- data.frame(bind_rows(df, preds))
    }
  }
  df = df[-1,]
  return(df)
}

#get daily centroid locations for all species in all years
dailylocs_east <- Estimatedailylocs(weight_yewa_east)
dailylocs_west <- Estimatedailylocs(weight_yewa_west)

#name color scale
cols <- c("1"="cadetblue", "2"="cadetblue", 
          "3"="darkolivegreen2", "4"="darkolivegreen3", "5"="darkolivegreen4", "6"="darkolivegreen", 
          "7"="darkorange1", "8"="darkorange2", "9"="darkorange3", "10"="darkorange4", 
          "11"="cadetblue", "12"="cadetblue")

ggplot() + 
  geom_polygon(data=coastlines, aes(x=long, y=lat, group = group), colour="grey50", fill="white") + 
  geom_point(data=dailylocs_east, aes(lon, lat, group=YEAR, col=as.factor(MONTH)), alpha=0.25, size=1) +
  geom_point(data=dailylocs_west, aes(lon, lat, group=YEAR, col=as.factor(MONTH)), alpha=0.25, size=1) +
 # geom_point(aes(col=winter), alpha=0.25) + 
  #scale_colour_viridis(discrete=TRUE) +
  scale_colour_manual(values = cols) +
  facet_wrap(~species) + 
  theme_void()
```

Plot latitudinal change
```{r}
ggplot(dailylocs, aes(DAY, lat)) + 
  geom_line(size=1) + #, aes(col=as.factor(MONTH))) + 
  #scale_colour_manual(values = cols) +
 # geom_line(data=dailylocs_west) + 
 # geom_line(data=dailylocs_east) +
  # geom_vline(xintercept=c(101, 209, 362), col="gray30") + #FIXME: Update xintercepts with dates
 xlab("Day of Year") + ylab("weighted mean latitude") +
  theme(element_text(size=20))+
  theme_bw()
```


Make a map comparing winter and summer locations #FIXME
```{r}
ggplot() + 
  geom_polygon(data=coastlines, aes(x=long, y=lat, group = group), colour="grey50", fill="white") + 
  #geom_polygon(data=countries, aes(x=long, y=lat, group=group), fill=NA, color="black")   +
  geom_polygon(data=grid, aes(x=long, y=lat, group=cell, fill=mean_weighted_count), size=0.001)    +
  geom_path   (data=grid, aes(x=long, y=lat, group=cell), alpha=0.4, color=NA) +
#  geom_point  (aes(x=cellcenters$lon_deg, y=cellcenters$lat_deg), size=0.5) +
  scale_fill_gradient(low="gray90", high="darkorchid4") + 
  #scale_fill_gradient2(low="blue", high="red", midpoint = 250) 
  theme_void() + 
  facet_wrap(~species)

```


For Figure 3:
```{r}
#count total number of records across the years. (increasing strongly)
counts <- robi_bchu_anhu %>%
  group_by(species, YEAR) %>%
  summarise(polyfid_num=n_distinct(POLYFID),
            total_num=sum(count.x), 
            total_eft=sum(count.y), 
            day_num=n_distinct(DAY), 
            prop_check=total_num/total_eft) %>% 
  filter(species %in% c("Archilochus alexandri", "Calypte anna"))

num_total <- ggplot(counts, aes(YEAR, total_num, group=species)) + 
  geom_line(aes(col=species)) + 
  geom_point(aes(col=species, shape=species))  +
  scale_color_viridis(discrete=TRUE, option="turbo") +
  scale_y_log10() +
  annotation_logticks (sides="l") + 
  xlab("Year") + 
  ylab("log10(Count of checklists)") +
  theme_bw() + 
  theme(text = element_text(size = 12))  + 
  scale_x_discrete(breaks=c("2008", "2010", "2012", "2014", "2016", "2018", "2020"))
  
num_checklists <- ggplot(counts, aes(YEAR, prop_check, group=species)) + 
  geom_line(aes(col=species)) + 
  geom_point(aes(col=species, shape=species))  +
  scale_color_viridis(discrete=TRUE, option="turbo") +
  xlab("Year") + 
  ylab("Percent of checklists") +
  theme_bw() + 
  theme(text = element_text(size = 12))  + 
  scale_x_discrete(breaks=c("2008", "2010", "2012", "2014", "2016", "2018", "2020"))

num_gridcells <- ggplot(counts, aes(YEAR, polyfid_num, group=species)) + 
  geom_line(aes(col=species)) + 
  geom_point(aes(col=species, shape=species)) +
  scale_color_viridis(discrete=TRUE, option="turbo") +
#  scale_y_log10() +
  xlab("Year") +
  ylab("Count of unique grid cells") + 
  theme_bw() + 
  theme(text = element_text(size=12))  + 
  scale_x_discrete(breaks=c("2008", "2010", "2012", "2014", "2016", "2018", "2020"))

num_daysofyear <- ggplot(counts, aes(YEAR, day_num, group=species)) + 
  geom_line(aes(col=species)) +
  geom_point(aes(col=species, shape=species)) +
  scale_color_viridis(discrete=TRUE, option="turbo") +
  xlab("Year") +
  ylab("Count of days observed") + 
  theme_bw() + 
  theme(text = element_text(size=12))  + 
  scale_x_discrete(breaks=c("2008", "2010", "2012", "2014", "2016", "2018", "2020"))

ggarrange(num_total, num_checklists, num_gridcells, num_daysofyear,
          labels=c("A", "B", "C", "D"),
          ncol=2, nrow=2, 
          legend="bottom",
          common.legend=TRUE)

ggsave(filename = "figs/ME_Fig3_2hbspecies_obs.png", height = 6, width=7)
```

