---
title: "process-eBird-data"
author: "Sarah Supp"
date: "2/19/2021"
output: html_document
---

##Code for cedar waxwing project, modified from eBird migration project (Supp et al. 2015)
(c) 2020, Niu and Supp
supps@denison.edu
Denison University
Code is under development, part of NSF Multi-Institution Collaborative Award (2019-22)

Birds evaluated include: 
* Cedar Waxwing (CEWA) _Bombycilla cedrorum_
* Robin (ROBI) _Turdus migratorius_
* Wood Thrush (WOTH) _Hylocichla mustelina_
* Yellow-rumped Warbler (YEWA) _setophaga coronata_
* Blue Jay (BLJA) _Cyanocitta cristata_


```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE)
```

If starting from scratch, input the three raw data files and merge together.
You need to bring in: 
1. a file with the bird observations (ebrd2.c)
2. a file with the total eBirder "effort" (eft.c)
3. a file with all the center locations for the POLYFID grid cells (locs)

Load the eBird effort and location files, from FAL (July 2020)
```{r}
# raw eBirder data, queried by FAL
 load("data/raw_eBird_data/effort2020.Rdata") # will come in as eft.c, FAL named previously
 load("data/raw_eBird_data/locs_2020.Rdata") # will come in as locs, FAL named previously
```

For *cedar waxwing*
```{r}
load("data/raw_eBird_data/Bombycilla_cedrorum.RData") # will come in as ebrd2.c, FAL named previously

# #merge locations data on POLYFID so it includes the center of each hex cell (LONGITUDE.y, LATITUDE.y)
  dat_merged = merge(ebrd2.c, locs, by.x = "POLYFID", by.y = "POLYFID")
  
  #now merge with the third data file for effort 
  #where count.x is number of target species and count.y is total number of eBird records (including those without the target species)
  dat_effort = merge(dat_merged, eft.c, by.x = c("POLYFID", "YEAR", "DAY"), by.y = c("POLYFID","YEAR",  "DAY")) 
  dat_effort$DAY <- as.numeric(dat_effort$DAY) 
  dat_effort$species <- "Bombycilla cedrorum"
  saveRDS(dat_effort, file = "data/dat_effort_CEWA.rds")
```

For *blue jay*
```{r}
load("data/raw_eBird_data/Cyanocitta_cristata.RData") # will come in as ebrd2.c, FAL named previously

# #merge locations data on POLYFID so it includes the center of each hex cell (LONGITUDE.y, LATITUDE.y)
  dat_merged = merge(ebrd2.c, locs, by.x = "POLYFID", by.y = "POLYFID")
  
  #now merge with the third data file for effort 
  #where count.x is number of target species and count.y is total number of eBird records (including those without the target species)
  dat_effort = merge(dat_merged, eft.c, by.x = c("POLYFID", "YEAR", "DAY"), by.y = c("POLYFID","YEAR",  "DAY")) 
  dat_effort$DAY <- as.numeric(dat_effort$DAY) 
  dat_effort$species <- "Cyanocitta cristata"
  saveRDS(dat_effort, file = "data/dat_effort_BLJA.rds")
```

For *wood thrush*
```{r}
load("data/raw_eBird_data/Hylocichla_mustelina.RData") # will come in as ebrd2.c, FAL named previously

# #merge locations data on POLYFID so it includes the center of each hex cell (LONGITUDE.y, LATITUDE.y)
  dat_merged = merge(ebrd2.c, locs, by.x = "POLYFID", by.y = "POLYFID")
  
  #now merge with the third data file for effort 
  #where count.x is number of target species and count.y is total number of eBird records (including those without the target species)
  dat_effort = merge(dat_merged, eft.c, by.x = c("POLYFID", "YEAR", "DAY"), by.y = c("POLYFID","YEAR",  "DAY")) 
  dat_effort$DAY <- as.numeric(dat_effort$DAY) 
  dat_effort$species <- "Hylocichla mustelina"
  saveRDS(dat_effort, file = "data/dat_effort_WOTH.rds")
```

For *robin*
```{r}
load("data/raw_eBird_data/Turdus_migratorius.Rdata") # will come in as ebrd2.c, FAL named previously

# #merge locations data on POLYFID so it includes the center of each hex cell (LONGITUDE.y, LATITUDE.y)
  dat_merged = merge(ebrd2.c, locs, by.x = "POLYFID", by.y = "POLYFID")
  
  #now merge with the third data file for effort 
  #where count.x is number of target species and count.y is total number of eBird records (including those without the target species)
  dat_effort = merge(dat_merged, eft.c, by.x = c("POLYFID", "YEAR", "DAY"), by.y = c("POLYFID","YEAR",  "DAY")) 
  dat_effort$DAY <- as.numeric(dat_effort$DAY) 
  dat_effort$species <- "Turdus migratorius"
  saveRDS(dat_effort, file = "data/dat_effort_ROBI.rds")
```


For *yellow-rumped warbler*
```{r}
load("data/raw_eBird_data/Setophaga_coronata.Rdata") # will come in as ebrd2.c, FAL named previously

# #merge locations data on POLYFID so it includes the center of each hex cell (LONGITUDE.y, LATITUDE.y)
  dat_merged = merge(ebrd2.c, locs, by.x = "POLYFID", by.y = "POLYFID")
  
  #now merge with the third data file for effort 
  #where count.x is number of target species and count.y is total number of eBird records (including those without the target species)
  dat_effort = merge(dat_merged, eft.c, by.x = c("POLYFID", "YEAR", "DAY"), by.y = c("POLYFID","YEAR",  "DAY")) 
  dat_effort$DAY <- as.numeric(dat_effort$DAY) 
  dat_effort$species <- "Setophaga coronata"
  saveRDS(dat_effort, file = "data/dat_effort_YEWA.rds")
```

