---
title: "seed-disperser-diversity"
author: "Sarah Supp"
date: "2025-03-03"
output: html_document
---

##Code for Eastern redcedar avian seed disperser project, modified from eBird project (Supp et al. 2015)
(c) 2025, Supp (PI) and La Sorte
supps@denison.edu
Denison University
Code is under development, part of NSF Multi-Institution Collaborative Award (2019-2025)

Birds evaluated include: 
* Cedar Waxwing (CEWA) _Bombycilla cedrorum_
* Robin (ROBI) _Turdus migratorius_
* Wood Thrush (WOTH) _Hylocichla mustelina_
* Yellow-rumped Warbler (YEWA) _setophaga coronata_
* Blue Jay (BLJA) _Cyanocitta cristata_
* European Starling (EUST) _Sturnus vulgaris_
* Eastern Bluebird (EABL) _Sialia sialis_
* Northern Mockingbird (NOMO) _Mimus polyglottos_
* Downy Woodpecker (DOWO) _Dryobates pubescens_
* Eastern Meadowlark (EAME) _Sturnella magna_
* White-breasted Nuthatch (WHNU) _Sitta carolinensis_
* Purple Finch (PUFI) _Haemorhous purpureus_
* Northern Cardinal (NOCA) _Cardinalis cardinalis_
* Dark-eyed Junco (DAJU) _Junco hyemalis_
* American Crow (AMCR) _Corvus brachyrhynchos_

The inputs to this code file are the results from a community-level analysis, shared by collaborator La Sorte. This includes for each hexagon cell (POLYFID) and year (16 winter seasons from 2008 to 2023) the number of target species (S; range=0-15), three diversity metrics (est; richness, shannon, simpson), and survey completeness estimates (sc). Values were calculated using the iNEXT package for the analysis and the same eBird data used in the population-level analysis. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
library(ggplot2)
library(ggridges)
```

Import the iNext data for the community analysis. It is based on the eBird occurrence information for all 15 seed dispersing species. Winter season was defined as November through February. We used years 2008-2023.
```{r}
# diversity estimates for winter seasons, by POLYFID hexagonal cell; loads 3 dataframes
load(here("data/1-raw_eBird_data/community/iNEXT-results.RData"))

# coordinates for the icosahedron cells
load(here("data/1-raw_eBird_data/locs.RData")) 
```

merge the diversity files with the location file so that polygon locations can be mapped
```{r}
# use merge to annotate SURVEY COMPLETENESS with the polygon coordinates (lat, long)
sc <- merge(sc, locs)

# use merge to annotate SPECIES RICHNESS with the polygon coordinates (lat, long)
S <- merge(S, locs)

# use merge to annotate DIVERSITY ESTIMATES with the polygon coordinates (lat, long)
est <- merge(est, locs)
```

Import polygons for plotting maps
```{r}
#Get polygons for the spatial range and make a map
world_map <- map_data("world")
north_america <- world_map %>%
  filter(region %in% c("USA", "Canada", "Mexico", "Guatemala", "Belize", 
                       "El Salvador", "Honduras", "Nicaragua", "Costa Rica", 
                       "Panama"))
```

# Survey completeness
Survey completeness of the bird occurrence information was estimated using the Hill numbers approach (Chao et al., 2014). We implement this analysis for each winter season using the iNEXT R package (Hsieh et al., 2024). We only included checklists in the survey completeness analysis where all the species detected by sight or sound were recorded by the observer (i.e. the ‘all observations reported’ field in eBird was designated as ‘yes’) (La Sorte et al., 2024). For our community analysis, we only examined spatial bins that were well surveyed in each season, with a survey completeness threshold >=90% (La Sorte et al., 2024).

Look at the survey completeness data, by year
```{r}
ggplot(data=sc, aes(x=SC, y=as.factor(year), group=as.factor(year), fill=year)) +
  geom_density_ridges() +
  scale_fill_viridis(name = "survey completeness", option = "C") +
  labs(y = "Winter Season Year",
       x = "Survey Completeness") +
  geom_vline(aes(xintercept=0.90)) +
  theme_ridges() +
    theme(
      legend.position="none",
      panel.spacing = unit(0.1, "lines"),
      strip.text.x = element_text(size = 8))
```
There is no major pattern in survey completeness across time, but maybe a slight increase in completeness in more recent years.


Check the relationship of observed species richness with survey completeness
```{r}
ggplot(sc, aes(S.obs, SC)) +
  geom_point(alpha=0.10) + 
  theme_bw()
```
As species richness increases, it is more likely that high sample completeness will be estimated. There are still a good nubmer of low species richness sites that also reach high survey completeness. 

Let's plot survey completeness on a map to show the estimates transparently
```{r}
#Prepare the grid cell data for mapping
# use the sepcies richness dataframe; and cell centers calculated
dgg <- dgconstruct(project = "FULLER", aperture = 4, topology = "HEXAGON", res = 6)
sc$cell <- dgGEO_to_SEQNUM(dgg, sc$x, sc$y)$seqnum
sc$cell_lat <- dgSEQNUM_to_GEO(dgg, sc$cell)$lat_deg 
sc$cell_lon <- dgSEQNUM_to_GEO(dgg, sc$cell)$lon_deg 

# Get the grid cell boundaries 
grid <- dggridR::dgcellstogrid(dgg, sc$POLYFID) 

#Update the grid cells' properties to include sc values
grid <- merge(grid, sc, by.x="seqnum", by.y="cell")

# #Make adjustments so the output is more visually interesting
grid <- grid %>% mutate(threshold=ifelse(SC<0.90, "exclude", "include"))
grid <- grid %>% mutate(thres_vals=ifelse(SC<0.90, NA, round(SC,2)))


# Winter survey completeness per grid cell (Nov-Feb)
ggplot() + 
  geom_sf(data = grid, aes(fill = thres_vals), color = alpha("white", 0.4 )) +
  geom_polygon(data=north_america, aes(x=long, y=lat, group=group), fill=NA, color="black")   +
  #scale_fill_discrete()+
  #scale_fill_gradient(low="grey90", high="black", na.value="white") +
  scale_fill_gradient(low="yellow", high="red") +
  coord_sf(xlim = c(-130, -60), ylim = c(10, 65)) + #set limits without distorting boundaries
  labs(title = "Winter survey completeness per grid cell (Nov-Feb)") +
  theme_minimal() + 
  theme(axis.line=element_blank(),axis.text.x=element_blank(),
          axis.text.y=element_blank(),axis.ticks=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank(),
          legend.position="bottom") +
  facet_wrap(~year, ncol=4)

ggsave(filename = here("figs/sample-coverage_map_08-23.png"), height = 11, width=8, bg="white")
```

#FIXME: Next step: For composition/diversity analysis, use the survey completeness threshold >=0.90 (90%) to filter the S and est dataframes. Cells <0.90 we cannot assume to be well sampled. 
- For each diversity estimate; calculate a linear regression, record slope, se, pvalue.
- should we also set a minimum number of years to be included to run the lm? 
- plot the results to view significant increases, significant decreases, and non-signficant cell trends (map, diverging histogram). Is there an overall direction of trend? Is there a spatial signal in the results? What different might we learn from S, Shan, Simp diversity? Is one of these better to pay attention to?
#FIXME: Also filter the cells to only include those within the eastern united states or within the ERC range: Need to add this step -- there are some south american cells, that aren't really relevant here. 






# Species richness
This file contains columns for POLYFID, year, and species richness. 
The 15 species listed at the top of this file were included; so species richness should range 0-15 for each POLYFID. 
What we want to know is: 
- Spatial: Are there areas with increased richness of ERC seed dispersers?
-- A hypothesis is that there should be higher richness of ERC seed dispersers in the southeastern united states, and in east coast urban areas (heat island, feeder effects, etc.)

- Temporal: Is seed disperser richness stable through time, or is there evidence of overall increase? 
--A hypothesis could be that changing climate is increasing seed disperser richness during the winter because of advanced migration and/or increased number of individuals or subpopulations that are not migrating southward, out of the ERC range.
--A spatial component could include a shift in areas that are experiencing this trend, and other areas that are not.

```{r}
ggplot(data=S, aes(x=S, y=as.factor(year), group=as.factor(year), fill=year)) +
  geom_density_ridges() +
  scale_fill_viridis(name = "species richness", option = "C") +
  labs(y = "Winter Season Year",
       x = "Species Richness") +
  theme_bw() +
    theme(
      legend.position="none",
      panel.spacing = unit(0.1, "lines"),
      strip.text.x = element_text(size = 8))
```

Show species richness on a map, separated by year
```{r}
#Prepare the grid cell data for mapping
# use the sepcies richness dataframe; and cell centers calculated
dgg <- dgconstruct(project = "FULLER", aperture = 4, topology = "HEXAGON", res = 6)
S$cell <- dgGEO_to_SEQNUM(dgg, S$x, S$y)$seqnum
S$cell_lat <- dgSEQNUM_to_GEO(dgg, S$cell)$lat_deg 
S$cell_lon <- dgSEQNUM_to_GEO(dgg, S$cell)$lon_deg 

# Get the grid cell boundaries 
grid <- dggridR::dgcellstogrid(dgg, S$POLYFID) 

#Update the grid cells' properties to include S values
grid <- merge(grid, S, by.x="seqnum", by.y="cell")


# Winter species richness per grid cell (Nov-Feb)
ggplot() + 
  geom_sf(data = grid, aes(fill = S), color = alpha("white", 0.4 )) +
  geom_polygon(data=north_america, aes(x=long, y=lat, group=group), fill=NA, color="black")   +
  #scale_fill_gradient(low="grey90", high="black", na.value="white") +
  scale_fill_gradient(low="yellow", high="red") +
  coord_sf(xlim = c(-130, -60), ylim = c(10, 65)) + #set limits without distorting boundaries
  labs(title = "Winter species richness per grid cell (Nov-Feb)") +
  theme_minimal() + 
  theme(axis.line=element_blank(),axis.text.x=element_blank(),
          axis.text.y=element_blank(),axis.ticks=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank(),
          legend.position="bottom") +
  facet_wrap(~year, ncol=4)

ggsave(filename = here("figs/S_map_08-23.png"), height = 11, width=8)

```

### Is there a significant trend for changing species richness across cells, through time?
Calculate the slopes of change across years, within cells. Plot the results to see if there is a trend of higher species disperser load through time. Need to think and check with FAL for ideas to see if we need to account for overall effort better (currently using as pres-only data, without weighted checks)
```{r}
# calculate separate lm for all grid cells
cells <- unique(S$POLYFID)

lm_S_results <- data.frame(cell=numeric(), n=integer(), slope=numeric(), 
                           pVal=numeric(), cell_lat=numeric(), cell_lon=numeric())

for (c in cells) {
  dat <- S %>% filter(POLYFID == c)
  #must have at least 3 data points to run a linear model; FIXME: set stronger requirement?
  if (nrow(dat) >= 3) {
      mod <- lm(S~as.numeric(year), data=dat)
      n = nrow(dat)
      slope = coef(mod)[[2]]
      pVal = glance(mod)$p.value[[1]]
      cell_lat = dat$cell_lat[1]
      cell_lon = dat$cell_lon[1]
      
      newrow <- data.frame(cell=c, n, slope, pVal, cell_lat, cell_lon)
      lm_S_results <- rbind(lm_S_results, newrow)
  }
}

#remove the NA values
lm_S_results <- na.omit(lm_S_results)
  
# add column for significance
lm_S_results$sig <- ifelse(lm_S_results$pVal <= 0.05, "sig", "NS")
```

Plot the results from the S linear regression
```{r}
#ggplot histogram, layered by identity for sig and NS
ggplot(lm_S_results, aes(slope)) + 
  geom_histogram(aes(fill=sig), binwidth=0.05, position="identity", alpha=0.5) +
  scale_fill_manual(values = c("gray40", "darkturquoise")) + 
  geom_vline(xintercept=0, linetype="dashed", col="gray10") +
  xlab("slope estimate") + 
  theme_bw()
```
Most of the significant slopes were positive (increasing richness). Very few significant negative slopes (decreasing richness).

```{r}
#average effect size, mean of the slopes
avg_effect_size <- lm_S_results %>% 
  filter(sig=="sig") %>%
  summarize(meanslope = mean(slope),
            medianslope = median(slope))

# is there any signal of the significant richness slopes by latitude (would indicate a N-S influence)
ggscatter(data=lm_S_results %>% filter(sig=="sig" & cell_lat>25), 
          x="cell_lat", y="slope", add="reg.line", alpha=0.25) +
  stat_cor(label.y = 0.8, aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~"))) +
  stat_regline_equation(label.y = 0.7) + 
  geom_hline(yintercept = 0, linetype="dashed") +
  theme(text = element_text(size=9), axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  #scale_x_continuous(breaks=seq(2008, 2024, 6)) +
  xlab("Cell Latitude") + ylab("S regression slope") + 
  theme_bw()
```

Let's plot the map using the significant slopes, and all others shaded in grey
```{r}
#Prepare the grid cell data for mapping
# use the sepcies richness dataframe; and cell centers calculated
dgg <- dgconstruct(project = "FULLER", aperture = 4, topology = "HEXAGON", res = 6)
lm_S_results$cell <- dgGEO_to_SEQNUM(dgg, lm_S_results$cell_lon, lm_S_results$cell_lat)$seqnum
lm_S_results$cell_lat <- dgSEQNUM_to_GEO(dgg, lm_S_results$cell)$lat_deg 
lm_S_results$cell_lon <- dgSEQNUM_to_GEO(dgg, lm_S_results$cell)$lon_deg 

# Get the grid cell boundaries 
grid <- dggridR::dgcellstogrid(dgg, lm_S_results$cell) 

#Update the grid cells' properties to include S values
grid <- merge(grid, lm_S_results, by.x="seqnum", by.y="cell")

# make subsets based on significant vs non-significant slope pvalues
gridsig <- grid %>% filter(sig == "sig")
gridNS <- grid %>% filter(sig == "NS")


# Winter species richness change (slope), per grid cell (Nov-Feb)
ggplot() + 
  geom_sf(data = gridsig, aes(fill = slope), color = alpha("white", 0.4 )) +
  geom_sf(data = gridNS, fill="grey", color = alpha("white", 0.04)) + 
  geom_polygon(data=north_america, aes(x=long, y=lat, group=group), fill=NA, color="black")   +
  #scale_fill_gradient(low="grey90", high="black", na.value="white") +
  scale_fill_gradient2(low="firebrick3", high="navyblue", midpoint = 0) +
  coord_sf(xlim = c(-130, -60), ylim = c(10, 65)) + #set limits without distorting boundaries
  labs(title = "Winter species richness slope per grid cell (2008-2023)") +
  theme_void() + 
  theme(axis.line=element_blank(),axis.text.x=element_blank(),
          axis.text.y=element_blank(),axis.ticks=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank(),
          legend.position="bottom")

ggsave(filename = here("figs/S_slope_map.png"), height = 4, width=6, bg="white")

```


