---
title: "ERC-range"
author: "Sarah Supp"
date: "10/8/2021"
output: html_document
---

##Code for cedar waxwing project, modified from eBird migration project (Supp et al. 2015)
(c) 2021, Yan and Supp
supps@denison.edu
Denison University
Code is under development, part of NSF Multi-Institution Collaborative Award (2019-2)

This document evaluates aspects of the *Eastern red cedar (Juniperus virginiana)* range.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(data.table)
library(devtools)
#devtools::install_github("dkahle/ggmap")
library(ggmap)
library(lubridate)
library(maptools)
library(fields)
library(sp)
library(raster)
library(rgdal)
library(rgeos)
library(sf)
library(maps)
library(mapdata)
library(rgdal)
library(tidyverse) 
library(ggpubr)
#devtools::install_github('r-barnes/dggridR', vignette=TRUE)
library(dggridR)
library(RColorBrewer)
library(geosphere)
library(radiant.data)
library(rnaturalearth)
#devtools::install_github("ropensci/rnaturalearthhires")
library(rnaturalearthhires)
#library(SDMTools)#FIXME: package ‘SDMTools’ is not available (for R version 3.6.2)
#devtools::install_github("r-spatial/sf") #C compiler error on mac
#library(geojsonio)
```



#Little's range and FIA importance value species distribution map for Eastern red cedar
Try something with a shapefile
Downloaded this one for eastern red cedar from https://www.fs.fed.us/nrs/atlas/littlefia/species_table.html  
```{r}
#found some useful help here: https://www.earthdatascience.org/courses/earth-analytics/spatial-data-r/intro-to-coordinate-reference-systems/ 
#may also be useful to read thorugh here: https://geocompr.robinlovelace.net/reproj-geo-data.html 

# read shapefile
ERC <- readOGR(dsn = paste0(here(), "/data/shapefile/litt68av.shp"))
```


```{r}
# convert to dataframe
ERC_df <- fortify(ERC)

# plot map using ggplot
ERCmap <- ggplot(ERC_df, aes(long,lat, group = group)) +
  geom_polygon(fill="darkolivegreen", col="black") +
  coord_equal() +
  labs(x = "Longitude (Degrees)",
       y = "Latitude (Degrees)",
      title = "ERC Range Map",
      subtitle = "D_Clarke_1866 Datum, Projection: Albers, Units: Degrees - Lat / Lon") + 
  theme_void()
ERCmap

# Figure for ESA 2021 poster
ggsave("ERC_range.png", ERCmap, bg="transparent")
```


TODO: FIXME: Need to figure out how to make this work so that it can be plotted onto a map of North America, and together with the eBird lat-long locations and icosahedron (hegaxon) pixels.
```{r}
#TODO reprojection of the range data is needed, or transformation of all other data including points to Albers, or it won't work.
# reproject data from Albers to WGS84 CRS
ERC_WGS84 = sf::st_transform(ERC, crs = "+proj=wgs84") #FIXME
ERC_WGS84 <- sp::spTransform(ERC,
                                CRS("+proj=wgs84")) #FIXME
```

FIXME: None of this works yet?
```{r}
# grab winter locations from the winter_dat_effort dataframe
# add a point to the map
mapLocations <- ERCmap +
                geom_point(data = winter_dat_effort,
                aes(x = lon, y = lat, group = NULL), colour = "orange",
                      size = .5)
mapLocations



--------------------------------
my_spdf <- readOGR( 
  dsn=paste0(here(), "/data/shapefile2/") , 
  layer="litt68av",
  verbose=FALSE
)

#explore the shapefile
summary(my_spdf)
length(my_spdf)
head(my_spdf@data)

# Basic plot of this shape file using Base R:
par(mar=c(0,0,0,0))
plot(my_spdf2, col="darkolivegreen", bg="grey80", lwd=0.25, border=0 )

# 'fortify' the data to get a dataframe format required by ggplot2
require(broom)
spdf_fortified <- tidy(my_spdf, region = "LITT68_ID")

# reproject data
spdf_WGS84 <- rgdal::spTransform(my_spdf,
                                crs(state_prov))

# get background map layers
state_prov <- rnaturalearth::ne_states(c("united states of america", "canada"))
us  <- map_data("usa")

# Plot it using ggplot2, just a shape, no context. Seems to be missing projection ?? Can't add points or background map.
ggplot() + 
  geom_polygon(data = spdf_fortified, aes( x = long, y = lat, group = group), fill="#69b3a2", color="white", alpha=0.25) +
  theme_void()

#TODO: Can't plot the shapefile onto the map because it isn't the right projection
#       sf should be abel to help this, maybe here? https://r-spatial.org/r/2018/10/25/ggplot2-sf-2.html 
ggplot(data = world) +
  geom_sf() + 
  geom_point(data=dat_effort, aes(x=lon, y=lat), size=0.2) + 
  coord_sf(xlim = c(-103, -48), ylim = c(24.5, 60), expand = FALSE) +
 #geom_polygon(data = spdf_fortified, aes( x = long, y = lat, group = group), fill="#69b3a2", color="white", alpha=0.25) +
  theme_void()
  
```


